---
title: "2023_Cerc_PDMR_analysis"
author: "Sam McNeill"
date: "2026-02-17"
output: word_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Load packages
```{r}
library(dplyr)
library(emmeans)
library(ggplot2)
library(lme4)
library(performance)
library(ARTool)
library(flextable)
library(viridis)

```

## Load and clean dataset for analysis and graphs

```{r}
# Read in data
data <- read.csv("data/2023_Cerc_Ratings.csv", header = T)
names <- read.csv("data/2023_Cerc_Names.csv", header = F)

# remove missing data and Pinnacle lines

names_to_remove <- c("Pc.1","Pc.2","Pc.3","Pc.4")

data_filtered <- data[!(data$Entry %in% names_to_remove), ]

data_clean <- data_filtered %>% left_join(names, by = c("Entry" = "V1"))

data_clean <- data_clean %>% filter(!is.na(Cerc.Rating.9.6.23))

colnames(data_clean)[16] <- c("Name")

# convert to factors
data_clean$Name <- as.factor(data_clean$Name)
data_clean$Rhizoc <- as.factor(data_clean$Rhizoc)
data_clean$Range <- as.factor(data_clean$Range)

# add ranks column
data_clean$ranks <- rank(data_clean$Cerc.Rating.9.6.23, ties.method = "average")


```

## Summary Stats

```{r}

colnames(names)[4] <- c("Name")


data_PI <- data %>% left_join(names, by = c("Entry" = "V1"))


sum_stats <- data_PI %>%
  group_by(Name) %>%
  summarise(
    "Total Observations" = n(),
    "Number Missing" = sum(is.na(Cerc.Rating.9.6.23)),
    "Number Rated" = n() - sum(is.na(Cerc.Rating.9.6.23)),
    "Median CLS Rating" = median(Cerc.Rating.9.6.23, na.rm = TRUE),
    "Min CLS Rating" = min(Cerc.Rating.9.6.23, na.rm = TRUE),
    "Max CLS Rating" = max(Cerc.Rating.9.6.23, na.rm = TRUE)
  )

sum_stats$`Min CLS Rating` <- ifelse(sum_stats$`Min CLS Rating` == Inf, "NA", sum_stats$`Min CLS Rating`)
sum_stats$`Max CLS Rating` <- ifelse(sum_stats$`Max CLS Rating` == -Inf, "NA", sum_stats$`Max CLS Rating`)

flextable(sum_stats) %>% 
  set_caption("Summary Stats of PIs at Final Rating") %>% 
  autofit() %>% 
  fit_to_width(max_width = 6.5)

```

## Heatmap of field

```{r}
ggplot(data_PI, aes(x = Row, y = Range, fill = Name)) +
  geom_tile(color = "white", size = 0.25) + 
  scale_fill_viridis_d() +
  labs(
    title = "Field Layout by Entry",
    x = "Row Number",
    y = "Range Number"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        legend.title = element_text(size = 8),
        legend.text  = element_text(size = 7),
        legend.key.size = unit(0.4, "cm"),
        legend.spacing.y = unit(0.1, "cm")
  )
```


```{r}
data_PI <- data_PI %>%
  mutate(highlight = ifelse(Entry == "CS", "CS", "Other"))

ggplot(data_PI, aes(x = Row, y = Range, fill = highlight)) +
  geom_tile(color = "white", size = 0.25) +
  scale_fill_manual(
    name = "Entry Type",
    values = c("CS" = "firebrick", "Other" = "grey80"),
    labels = c("Commercial Susceptible (CS)", "Other Entries")
  ) +
  labs(
    title = "Field Layout Highlighting CS Entries",
    subtitle = "CS entries shown in red",
    x = "Row Number",
    y = "Range Number"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "bottom"
  )
```


## Boxplot with missing data removed

```{r}
ggplot(data_clean, aes(x = reorder(Name, Cerc.Rating.9.6.23, FUN = median),
                       y = Cerc.Rating.9.6.23)) +
  geom_boxplot() +
  coord_flip() + 
  labs(x = "PI Name", y = "Disease Rating (0-10)", title = "Cercospora Leaf Spot Disease Ratings by PI (missing data excluded)") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))


```

## Test to see if covariates are meaningful
```{r}
linear_model <- lm(ranks ~ Name + Rhizoc + Range, data = data_clean)
summary(linear_model)

mixed_model <- lmer(ranks ~ Name + (1|Rhizoc) + (1|Range), data = data_clean)

#summary(mixed_model)

icc_value <- icc(mixed_model)
print(icc_value)

```

Testing Kruskal-Wallis and Dunns
```{r}
library(rstatix)

kruskal.test(Cerc.Rating.9.6.23 ~ Name, data = data_clean)

data_clean %>% dunn_test(Cerc.Rating.9.6.23 ~ Name, p.adjust.method = "holm") %>% filter(grepl("EL50/2", group1))
```

KW and Dunns don't give enough power. However, Rhizoc and Range are statistically significant at a = 0.05 in the parametric LM but this assumes normality. An ICC of 40% indicates substantial clustering. This justifies a mixed model framework with random effects rather than a simple nonparametric test like Dunn's.


## Will use ART to accommodate the random effects while maintaining nonparametric framework

Does Rhizoc as a whole affect ranked disease ratings?

Rhizoc only has two levels, which is insufficient for variance estimation (random effect). We will include it as a fixed interaction (interaction required for ART fixed effects) and look at significance.
```{r}
art_model <- art(Cerc.Rating.9.6.23 ~ Name * Rhizoc + (1|Range), data = data_clean)

summary(art_model)
anova(art_model)
```
Rhizoc and interaction coeffs are not significant so we drop Rhizoc for simplicity. Range has 22 levels so it will be included as random effect to account for blocking. 

```{r}
# Simpler model without Rhizoc
art_model_simple <- art(Cerc.Rating.9.6.23 ~ Name + (1|Range), 
                        data = data_clean)
anova(art_model_simple)
```
Lines have significant differences, let's check each year and compare to resistant controls

Run ART model for EL50/2 (res check) for all time points
```{r}
# set reference level for dunnetts comparison
ref_level <- "EL50/2"
data_clean$Name <- relevel(data_clean$Name, ref = ref_level)

# initalize dates and functions
rating_dates <- c("Cerc.Rating.8.16.23", "Cerc.Rating.8.21.23", 
                  "Cerc.Rating.8.28.23", "Cerc.Rating.9.6.23")

date_labels <- c("16-Aug", "21-Aug", "28-Aug", "6-Sep")

get_stars <- function(p) {
  case_when(
    p < 0.001 ~ "***",
    p < 0.01 ~ "**",
    p < 0.05 ~ "*",
    p < 0.1 ~ ".",
    TRUE ~ ""
  )
}

# Run ART for each date and extract median
all_results <- lapply(seq_along(rating_dates), function(i) {
  date <- rating_dates[i]
  label <- date_labels[i]
  
  # Fit ART model
  formula <- as.formula(paste(date, "~ Name + (1|Range)"))
  model <- art(formula, data = data_clean)
  
  # Contrasts vs Res control
  contrasts_all <- as.data.frame(summary(art.con(model, "Name", adjust = "dunnett")))
  
  contrasts <- summary(contrasts_all) %>% 
    filter(grepl(ref_level, contrast))
  
  # Get medians
  medians <- data_clean %>%
    group_by(Name) %>%
    summarise(median = median(.data[[date]], na.rm = TRUE))
  
  ref_pattern <- paste0("^\\(", gsub("([/\\-])", "\\\\\\1", ref_level), "\\) - ")
  
  # Extract stars for each entry from contrast output
  stars <- contrasts %>%
    mutate(Name = gsub(ref_pattern, "", contrast), # remove ref to get just line name
           Name = gsub("^\\(", "", Name),   # remove leading (
           Name = gsub("\\)$", "", Name),   # remove trailing )
           Name = trimws(Name), # trim whitespace
           stars = get_stars(p.value)) %>%
    select(Name, stars)
  
  # Combine median and stars
  medians %>%
    left_join(stars, by = "Name") %>%
    mutate(stars = ifelse(is.na(stars), "", stars),
      !!label := paste0(median, stars)) %>%
    select(Name, !!label)
  
})

```


Table with median dates and significance
```{r}
# Join all dates into one table
final_table <- Reduce(function(x,y) left_join(x, y, by = "Name"), all_results)

# Print table
flextable(final_table) %>%
  set_caption("Cercospora Leaf Spot Median Ratings by Entry and Date") %>%
  autofit()
```





```{r}
library(grateful)
cite_packages(output = "paragraph", out.dir = ".")
```

Wobbrock, J. O., Findlater, L., Gergle, D., and Higgins, J. J. 2011. The Aligned Rank Transform for Nonparametric Factorial Analyses Using Only ANOVA Procedures. Pages 143-146 in: Proceedings of the ACM Conference on Human Factors in Computing Systems (CHI '11). Association for Computing Machinery, New York, NY. https://doi.org/10.1145/1978942.1978963